Thesis: draft

### 2021-05-29 draft.tex
latest update

\addcontentsline{toc}{chapter}{Abstract}
\noindent
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
\newline
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effect of algorithmic personalization on the formation of phenomena such as polarization, spread of extremism and formation of echo chambers.\\
\newline
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their opinion. This aspect of potential stubbornness should not be ignored.\\
\newline
In this thesis toy models are designed to study the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions. We will investigate whether network properties such as community structure and clustering may enhance or hamper the formation of echo chambers. We will use filtering algorithms to determine their possible effects on the formation of echo chambers and we will study the impact of individual stubbornness. We will also study different ways of distributing the opinions and how this may affect the opinion dynamics in the system. \\
\newline
Finally, we will test the model on real-world networks. We will investigate whether the results agree with results from the theoretical network models.

@@

Sociale media zijn bijna niet meer weg te denken uit ons dagelijks leven. De laatste tientallen jaren is het gebruik van sociale media, zowel op persoonlijk vlak als op professioneel vlak, in snel tempo toegenomen. Tegenwoordig kan men zeggen dat (bijna) iedereen, naast zijn werkelijke leven, een virtueel leven op sociale media leidt. Het gebruik van sociale media heeft ongetwijfeld veel voordelen. Goed en kwaad gaan echter vaak hand in hand en men kan niet ontkennen dat het gebruik van sociale media ook veel potentiële, soms ongekende, nadelen kent. Eén van die, tot nog toe, ongekende kanten van sociale media is de impact op de dynamica van opinies in de maatschappij. \\

@@

Het is gekend dat mensen niet enkel hun mening uit zichzelf vormen, maar sterk beïnvloed worden door onder andere de mensen rond hen, hun omgeving en hun opvoeding. Sociale media laten onmiddellijke interacties toe tussen mensen van overal op de wereld, zonder daarbij belemmerd te worden door fysieke- en tijdsgrenzen. Dit kan een potentieel grote impact hebben op hoe meningen zich vormen en hoe ze evolueren.\\

@@

Door het toenemende gebruik worden sociale media echter overspoeld door informatie. Om hun gebruikers niet het gevoel te geven dat ze verdrinken in wat ze te zien krijgen op hun sociale media platform, gebruiken deze sociale media platformen filtering algoritmen. Deze filtering algoritmen filteren, zoals de naam al impliceert, en individualiseren de informatie die iemand te zien krijgt op zijn of haar tijdslijn. Een gebruiker op sociale media krijgt dus niet alle berichten die zijn of haar vrienden posten te zien, maar enkel een selectie die onder andere gemaakt wordt op basis van de gebruiker zijn of haar sociale media geschiedenis. Dit filteren en individualiseren kan, mogelijks, grote gevolgen hebben. Zo zou het kunnen leiden tot de vorming van ``echo kamers'' waar gebruikers gevangen raken tussen ideeën die hun eigen visie versterken. Dit kan op zijn beurt polarisatie in de maatschappij, opkomst van extremisme, enzovoort bevorderen. Het is daarom uiterst belangrijk om de impact van deze filtering algoritmen grondig in kaart te brengen.\\

@@

In deze thesis wordt gebruik gemaakt van ``toy models'' die de interactie tussen verschillende netwerk modellen, filtering algoritmen en individuele terughoudenheid om van mening te veranderen onderzoeken in de evolutie van twee opinies. Er worden verschillende netwerk modellen geïntroduceerd die de impact van onder andere clustering en groepsstructuur op de evolutie van opinies en de vorming van echo kamers moeten onderzoeken. Daarnaast wordt er gebruik gemaakt van enkel simpele filtering algoritmen om hun impact op de dynamica van de opinies te onderzoeken. Tot slot wordt er een notie van koppigheid in het systeem geïntroduceerd om mogelijke effecten hiervan in kaart te kunnen brengen. \\

@@

Naast het bovenstaande zal er ook kort aandacht gespendeerd worden aan hoe de twee verschillende opinies verdeeld worden over het systeem. We zullen de opinies in groepsstructuren verdelen en dit scenario vergelijken met het scenario waarin de opinies willekeurig verdeeld worden. Om de thesis te besluiten zal het opinie-dynamica model, dat in de thesis gebruikt werd, getest worden op netwerken verkregen van data uit de werkelijke wereld (bv. van email data of telefoonnetwerken). Er zal nagegaan worden of de resultaten van deze werkelijke netwerken gelijkenissen, dan wel verschillen, vertonen met de data verkregen van de netwerk modellen.

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have an impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
Some people change their opinions more easily than others, that is, once in a while you encounter somebody that is stubborn and persistent towards its own opinion. It is important to understand the effects of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant and convenient experience on their platforms, social media companies use algorithms to order and filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of importance to know the effect of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate models to reproduce the opinion dynamics and opinion formation processes found in real life.\\
\newline
Section \ref{statPhys} introduces concepts such as statistical and social physics. Section \ref{complNet} gives an introduction to the first layer, that of complex networks. The second layer, the opinion dynamics, is introduced in Section \ref{Opinion}. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

An important tool in the study of opinion dynamics in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms and molecules to macroscopic observed behavior and has a wide range of applications inside and outside physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics \cite{Castellano2009}\cite{Galam2008}\cite{Galam1982}\cite{Stauffer2012}. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display regularities, transitions from disorder to order, the emergence of consensus, universality \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the regularities at large scale as collective effects of the interactions among individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one difficulty compared to the description of gases and other physical phenomena. In physics, the microscopic constituents are atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics of individuals would be well know, the interactions between them would still be complicated and it would be hard to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, extracting useful and qualitative information out of social models is a challenging task \cite{Castellano2009}. The use of statistical physics may, however, offer a way out. In statistical physics it is found that different systems behave qualitatively the same and are governed by the same laws near the critical point \cite{Hu2018}\cite{Kadanoff2010}. This lead to the concept of universality: many different microscopic problems have the same macroscopic manifestation \cite{Gug2015}\cite{Kadanoff2015}. This idea may then be extended to social systems \cite{Hu2018}: we should not take the unique character of every individual into account, instead these individual characteristics can be incorporated in a sort of averaged way \cite{Gug2015}. One can then proceed to build models of social behavior where only the most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with empirical data to investigate whether the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in physical research and science in general.

@@

Complex networks have become more and more popular for analyzing complex, dynamical systems. They are used in fields such as physics, economics, social sciences and biology \cite{Costa2008}. These different fields are diverse, but they have at least one common ground: they often deal with a large number of components that interact with each other. In other words, in all these fields one encounters complex systems, systems where it is not possible to predict collective behavior based on the properties of the individual components alone \cite{Mata2020}. Complex systems often display phenomena such as non-linearity, emergence and spontaneous order.
\\
\newline
One of the tools to deal with these complex systems and to give us more insight in the possible underlying structures are complex networks. A network is a structure composed of nodes or vertices and a set of links or edges that indicate the interactions between the nodes (in this thesis nodes will mostly be called actors; edges will be called edges) \cite{Mata2020}.\\
\newline
Representing a complex system as a network makes the system appear simpler and tractable, while it still includes the non-linearity of these systems. One can find the language to describe networks in mathematical graph theory. However, complex systems in real life often deal with a large number of components, so the use of statistical and high-performance computing tools is inevitable \cite{Mata2020}. One could argue that the study of complex networks lies somewhere at the intersection of mathematical graph theory and statistical physics \cite{F.Costa2007}.\\
\newline
The advantage of working with network models is that they reduce the level of complexity encountered in the real world, so that one can treat these systems in a more practical way. However, we still want our models to display properties similar to the ones seen in real systems \cite{Mata2020}. Since many real systems are not static but evolve in time, this means that we do not only need to deal with static networks but also with temporal networks. Temporal networks are networks where the edges are not always active, but instead become active for some periods of time \cite{Holme2012}. In temporal networks time becomes an explicit element in the network representation \cite{Holme2012}. Static networks have been widely studied and are often convenient for their analytic tractability, whereas temporal networks are, in some cases, more realistic \cite{Mata2020}.\\
\newline
Some other important concepts in network theory are the degree distribution, clustering, community structure and connectivity. These concepts will be explained in depth in Section \ref{netDef}, but let us already anticipate on the case of social networks. Many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering, also referred to as ``small world phenomenon'' (note that in the literature the term ``small world'' sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of them. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\
\newline
The theoretical network models used in this thesis are the Erd\H{o}s-R\'{e}nyi network, the stochastic block model and the Watts-Strogatz model. These will be explained thoroughly in Section \ref{netModel}.

@@

The formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (e.g. stubbornness) and the information someone is exposed to. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or actors in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization and formation of echo chambers (clusters of people with the same opinion) \cite{Sirbu2016}. \\
\newline
The opinion of the actors in the model can be either discrete or continuous. This thesis deals with models where each actor can have one of two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (e.g. republicans or democrats, renting or buying a house). \\
\newline
The rules that determine how an actor updates his or her opinion can include different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a probabilistic way of updating. An overview of possible opinion dynamics models is given in Section \ref{Bin} and Section \ref{Con}. \\
\newline
One can also include a resistance parameter that determines the hesitation of an actor to change to a new opinion or a parameter that determines the influence of an actor on others. Furthermore, the actors can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Section \ref{stubb}; the activation mechanism of actors is discussed in Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define different filtering algorithms. The filtering algorithms that social media companies use, are corporate secrets, but they typically have three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Section \ref{filter} and Section \ref{modelThesis} a detailed description of the filtering algorithms used in this thesis will be given.
\newpage
\section{This thesis: hypotheses and goals}\label{goal}

This thesis will study the interplay between social network structure and individual resistance to change in the evolution of two competing opinions. On one hand, we will determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we will study the effect of stubborn actors in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms will be included.\\ 
\newline
Perra and Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions \cite{Perra2019}. They also considered the effect of nudging (meaning that one opinion is pushed to all actors in the network). Their main findings were that algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering) could result in the formation of echo chambers. On the other hand, a more heterogeneous contact pattern hampers this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}. They, however, did not study the effect of stubborn actors in the network, nor did they investigate networks with community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have community structure and relatively low clustering to networks without community structure, but with high clustering (a detailed description of concepts like clustering and community structure will be given in Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with high clustering. We will also study networks that have both high clustering and community structure to identify the conditions and driving forces leading to the formation of echo chambers.\\
\newline
The impact of stubborn actors on the formation and evolution of opinions in the network will be studied. We aim to determine whether there is an interplay between network structure, filtering algorithm and stubbornness.

@@

$N_{\Delta}(i)$ counts the number of edges between the neighbors of $i$ and $N_3(i)$ is equal to $k_i(k_i-1)/2$, where $k_i$ is the degree of node $i$. The local clustering coefficient $cc_i$ (Eq. (\ref{clus})) of a node and the transitivity $T_i$ (Eq. (\ref{trans})) of a node define the same quantity \cite{F.Costa2007}. This is not the case for the average clustering coefficient $\left<cc\right>$ and the global transitivity $T$. The difference between the two definitions is that the average of Eq. (\ref{globalTrans}) gives the same weight to each triangle in the network, whereas Eq. (\ref{avClus}) gives the same weight to each node \cite{Newman2003}. This may lead to slightly different values since nodes with a higher degree may possibly be involved in a higher number of triangles than nodes with a lower degree \cite{F.Costa2007}.

@@

The modularity of a network with $N$ nodes and $L$ edges and partitioned into $n_c$ communities is calculated as follows. Each community has $N_c$ nodes and $L_c$ edges ($c = 1,..., n_c$) \cite{Albert2016}. The modularity of a community $Q_c$ is then calculated by measuring the difference between the network's real wiring diagram (which is given by the adjacency matrix $A_{ij}$) and the expected number of edges between $i$ and $j$ if the network is randomly wired (where $i$ and $j$ are two nodes that lie in the same community $C_c$) \cite{Albert2016}. This last quantity is given by $p_{ij}$, which is obtained by randomizing the original network while keeping the expected degree of each node unchanged \cite{Albert2016}. Hence, the modularity of a community $Q_c$ becomes

@@

The derivations of these formulas can be found in Appendix \ref{WS}, Section \ref{ringclus}. The regular ring lattice thus has a high clustering coefficient and an average path length that grows linearly with the system size $N$. \\

@@

\textbf{The stochastic Watts-Strogatz block model (SBM-WS)}\\

>>>>

\addcontentsline{toc}{chapter}{Abstract}
\noindent
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
\newline
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bound by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downsides in the sense that people might get trapped in ``echo chambers'' that confirm their own beliefs. It is important to know the effect of algorithmic personalization on phenomena such as polarization, spread of extremism and formation of echo chambers.\\
\newline
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their opinion. This aspect of potential stubbornness should not be ignored.\\
\newline
In this thesis toy models are designed to study the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions. We will investigate whether network properties such as community structure and clustering may enhance or hamper the formation of echo chambers. We will use filtering algorithms to determine their possible effect on the formation of echo chambers and we will study the impact of individual stubbornness. We will also study different ways of distributing the opinions and how this may affect the opinion dynamics. \\
\newline
Finally, we will test the model on real-world networks. We will investigate whether the results agree with results from theoretical network models.

@@

Sociale media zijn niet meer weg te denken uit ons dagelijks leven. De laatste tientallen jaren is het gebruik van sociale media, zowel op persoonlijk vlak als op professioneel vlak, in snel tempo toegenomen. Tegenwoordig kan men zeggen dat (bijna) iedereen, naast zijn werkelijke leven, een virtueel leven op sociale media leidt. Het gebruik van sociale media heeft ongetwijfeld veel voordelen. Goed en kwaad gaan echter vaak hand in hand en men kan niet ontkennen dat het gebruik van sociale media ook veel potentiële, vaak ongekende, nadelen kent. Eén van die, tot nog toe, ongekende kanten van sociale media is de impact op de dynamica van opinies in de maatschappij. \\

@@

Het is gekend dat mensen niet enkel hun mening uit zichzelf vormen, maar beïnvloed worden door onder andere de mensen rond hen, hun omgeving en hun opvoeding. Sociale media laten onmiddellijke interacties toe tussen mensen van overal op de wereld, zonder daarbij belemmerd te worden door fysieke- en tijdsgrenzen. Dit kan een grote impact hebben op hoe meningen zich vormen en hoe ze evolueren.\\

@@

Door het toenemende gebruik worden sociale media echter overspoeld door informatie. Om hun gebruikers niet het gevoel te geven dat ze verdrinken in wat ze te zien krijgen op hun sociale media platform, gebruiken deze sociale media platformen filtering algoritmen. Deze filtering algoritmen filteren, zoals de naam al impliceert, en individualiseren de informatie die iemand te zien krijgt op zijn of haar tijdslijn. Een gebruiker op sociale media krijgt dus niet alle berichten die zijn of haar vrienden posten te zien, maar enkel een selectie die onder andere gemaakt wordt op basis van de gebruiker zijn of haar sociale media geschiedenis. Dit filteren en individualiseren kan, mogelijks, grote gevolgen hebben. Zo zou het kunnen leiden tot de vorming van ``echo kamers'' waar gebruikers gevangen raken tussen ideeën die hun eigen visie versterken. Dit kan op zijn beurt polarisatie in de maatschappij, opkomst van extremisme, enzovoort bevorderen. Het is daarom belangrijk om de impact van deze filtering algoritmen grondig in kaart te brengen.\\

@@

In deze thesis wordt gebruik gemaakt van ``toy models'' die de interactie tussen verschillende netwerk modellen, filtering algoritmen en individuele terughoudenheid om van mening te veranderen onderzoeken in de evolutie van twee opinies. Er worden verschillende netwerk modellen geïntroduceerd die de impact van onder andere clustering en groepsstructuur op de evolutie van opinies en de vorming van echo kamers moeten onderzoeken. Daarnaast wordt er gebruik gemaakt van simpele filtering algoritmen om hun impact op de dynamica van de opinies te onderzoeken. Verder wordt er een notie van koppigheid geïntroduceerd om mogelijke effecten hiervan in kaart te kunnen brengen. \\

@@

Naast het bovenstaande zal er ook kort aandacht gespendeerd worden aan hoe de twee verschillende opinies verdeeld worden over het systeem. We zullen de opinies in groepsstructuren verdelen en dit scenario vergelijken met het scenario waarin de opinies willekeurig verdeeld worden. Om de thesis te besluiten zal het opinie-dynamica model, dat in de thesis gebruikt wordt, getest worden op netwerken verkregen van data uit de werkelijke wereld (bv. van email data of telefoonnetwerken). Er zal nagegaan worden of de resultaten van deze werkelijke netwerken gelijkenissen, dan wel verschillen, vertonen met de data verkregen van netwerk modellen.

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have an impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
\newline
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
Some people change their opinions more easily than others, that is, once in a while you encounter somebody that is stubborn and persistent towards its own opinion. It is important to understand the effect of individual resistance on the formation and evolution of opinions. When designing models of opinion dynamics on social networks, this individual stubbornness should not be ignored.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant and convenient experience on their platforms, social media companies use algorithms to order and filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of importance to know the effect of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate models to reproduce the opinion dynamics and opinion formation processes found in real life.
\newpage
\noindent
Section \ref{statPhys} introduces concepts such as statistical and social physics. Section \ref{complNet} gives an introduction to the first layer, that of complex networks. The second layer, the opinion dynamics, is introduced in Section \ref{Opinion}. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are summarized and discussed.

@@

An important tool in the study of opinion dynamics in large groups of people is the use of statistical physics. This branch of physics offers a framework that relates microscopic properties of atoms and molecules to macroscopic observed behavior and has a wide range of applications inside and outside physics. The observation that large number of people display collective, `macroscopic' behavior begged for the use of the concepts and insights developed in statistical physics \cite{Sirbu2016}. The application of the theory of statistical physics on social phenomena is referred to as social physics \cite{Castellano2009}\cite{Galam2008}\cite{Galam1982}\cite{Stauffer2012}. The `microscopic' constituents are now individual humans who interact with a limited number of other individuals and in that way form complex, `macroscopic' groups such as human societies. These `macroscopic' groups display regularities, transitions from disorder to order, the emergence of consensus, universality \cite{Buchanan2007}. The statistical physics approach of these social systems tries to explain the regularities at large scale as collective effects of the interactions among individuals \cite{Sirbu2016}.\\
\newline
The statistical approach of social phenomena has one difficulty compared to the description of gases and other physical phenomena. In physics, the microscopic constituents are atoms or molecules, whose dynamics and interactions with each other are often well known and can be put in relatively simple laws with only a few parameters \cite{Castellano2009}. Humans, on the other hand, are far from simple entities and the dynamics of a single individual are far from known \cite{Castellano2009}. Even if the dynamics of individuals would be well know, the interactions between them would still be complicated and it would be hard to describe them with simple laws and only a few parameters \cite{Castellano2009}. Hence, extracting useful and qualitative information out of social models is a challenging task \cite{Castellano2009}. The use of statistical physics may, however, offer a way out. In statistical physics it is found that different systems behave qualitatively the same and are governed by the same laws near the critical point \cite{Hu2018}\cite{Kadanoff2010}. This lead to the concept of universality: many different microscopic problems have the same macroscopic manifestation \cite{Gug2015}\cite{Kadanoff2015}. This idea may be extended to social systems \cite{Hu2018}: we should not take the unique character of every individual into account, instead these individual characteristics can be incorporated in a sort of averaged way \cite{Gug2015}. One can then proceed to build models of social behavior where only the most important properties of single individuals and their interactions are taken into account \cite{Castellano2009}. An important factor herein is the comparison with empirical data to investigate whether the trends seen in real life are compatible with the outputs generated by the models \cite{Castellano2009}\cite{Pawel}. This interplay between theory, model, experiment and real-life data is not only an important factor in social physics, but also in physical research and science in general.

@@

Complex networks have become more and more popular for analyzing complex, dynamical systems. They are used in fields such as physics, economics, social sciences and biology \cite{Costa2008}. These different fields are diverse, but they have at least one common ground: they often deal with a large number of components that interact with each other. In other words, in all these fields one encounters complex systems, systems where it is not possible to predict collective behavior based on the properties of the individual components alone \cite{Mata2020}. Complex systems often display phenomena such as non-linearity, emergence and spontaneous order.
\\
\newline
One of the tools to deal with these complex systems and to give us more insight in the underlying structures are complex networks. A network is a structure composed of nodes or vertices and a set of links or edges that indicate the interactions between the nodes (in this thesis nodes will mostly be called actors; edges will be called edges) \cite{Mata2020}.\\
\newline
Representing a complex system as a network makes the system appear simpler and tractable, while it still includes the non-linearity of these systems. One can find the language to describe networks in mathematical graph theory. However, complex systems in real life often deal with a large number of components, so the use of statistical and high-performance computing tools is inevitable \cite{Mata2020}. One could argue that the study of complex networks lies somewhere at the intersection of mathematical graph theory and statistical physics \cite{F.Costa2007}.\\
\newline
The advantage of working with network models is that they reduce the level of complexity encountered in the real world, so that one can treat these systems in a more practical way. However, we still want our models to display properties similar to the ones seen in real systems \cite{Mata2020}. Since many real systems are not static but evolve in time, this means that we do not only need to deal with static networks but also with temporal networks. Temporal networks are networks where the edges are not always active, but instead become active for some periods of time \cite{Holme2012}. Hence, time becomes an explicit element in the network representation \cite{Holme2012}. Static networks have been widely studied and are often convenient for their analytic tractability, whereas temporal networks are, in some cases, more realistic \cite{Mata2020}.\\
\newline
Some other important concepts in network theory are the degree distribution, clustering, community structure and connectivity. These concepts will be explained in depth in Section \ref{netDef}, but let us already anticipate on the case of social networks. Many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering, also referred to as ``small world phenomenon'' (in the literature the term ``small world'' sometimes only refers to the property of short distances; in this thesis, it is used to indicate both high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of them. They might, however, still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\
\newline
The theoretical network models used in this thesis are the Erd\H{o}s-R\'{e}nyi network, the stochastic block model and the Watts-Strogatz model. These will be explained in Section \ref{netModel}.

@@

The formation of an individual's opinion is the result of the interplay of many factors such as self-reflection, peer pressure, the individual's personality (e.g. stubbornness) and the information someone is exposed to. Opinion dynamics models should try to capture these complex processes in a simplified way. Several models have been developed ranging from simple binary models to more complex, continuous approaches \cite{Sirbu2016}.\\
\newline
The basic idea of all opinion dynamics models is that the nodes or actors in a social network have a variable that represents their opinion and that is updated according to some predefined rules. These models are a simplification of real-world opinion dynamics. It is however shown that they display aspects of real opinion formation such as agreement, transitions between order (consensus) and disorder (fragmentation), polarization and formation of echo chambers (clusters of people with the same opinion) \cite{Sirbu2016}. \\
\newline
The opinion of the actors in the model can be either discrete or continuous. This thesis deals with models where each actor can have one of two opinions, A or B, and thus only deals with the case of discrete opinions. This might come across as a simplification of the real-world complexity of opinion formation, but also in real life people often have to choose between two competing opinions (e.g. republicans or democrats, renting or buying a house). \\
\newline
The rules that determine how an actor updates his or her opinion can include different aspects. They can, for example, be as simple as a majority rule (i.e. the majority model: if the majority of your neighbors have a certain opinion, you adopt that opinion as well) or can include a probabilistic way of updating. An overview of possible opinion dynamics models is given in Section \ref{Bin} and Section \ref{Con}. One can also include a resistance parameter that determines the hesitation of an actor to change to a new opinion or a parameter that determines the influence of an actor on others. Furthermore, the actors can have a parameter that determines whether they are active or not. The introduction of stubborn actors will be discussed in Section \ref{stubb}; the activation mechanism of actors is discussed in Section \ref{actMech}.\\
\newline
Since this thesis deals with the particular case of opinion dynamics on on-line social platforms, it is important to define filtering algorithms. The filtering algorithms that social media companies use, are corporate secrets, but they typically have three main principles of content curation: popularity, semantic and collaborative filtering \cite{Perra2019}. Popularity filtering refers to the practice of promoting content that is popular across the platform; semantic filtering means that post similar to previous consumed posts are recommended and collaborative filtering suggests posts that are similar to the ones our friends consume \cite{Perra2019}. In Section \ref{filter} and Section \ref{modelThesis} a detailed description of the filtering algorithms used in this thesis will be given.
\newpage
\section{This thesis: hypotheses and goals}\label{goal}

This thesis will study the interplay between social network structure and individual resistance to change in the evolution of two competing opinions. On one hand, we will determine the influence of network structures on the formation and evolution of opinions, whereas, on the other hand, we will study the effect of stubborn actors in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms will be included.\\ 
\newline
Perra and Rocha investigated both the effect of filtering algorithms and network topologies on the formation and evolution of two competing opinions \cite{Perra2019}. They also considered the impact of nudging (meaning that one opinion is pushed to all actors in the network). Their main findings were that algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering) could result in the formation of echo chambers. On the other hand, a more heterogeneous contact pattern hampers this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}. They, however, did not study the effect of stubborn actors in the network, nor did they investigate networks with community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have community structure and relatively low clustering to networks without community structure, but with high clustering (a detailed description of concepts like clustering and community structure will be given in Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with high clustering. We will also study networks that have both high clustering and community structure to identify the conditions and driving forces leading to the formation of echo chambers.\\
\newline
The impact of stubborn actors on the formation and evolution of opinions in the network will be studied. We aim to determine whether there is an interplay between network structure, filtering algorithm and stubbornness.

@@

$N_{\Delta}(i)$ counts the number of edges between the neighbors of $i$ and $N_3(i)$ is equal to $k_i(k_i-1)/2$, where $k_i$ is the degree of node $i$. The local clustering coefficient $cc_i$ (Eq. (\ref{clus})) and the transitivity $T_i$ (Eq. (\ref{trans})) of a node define the same quantity \cite{F.Costa2007}. This is not the case for the average clustering coefficient $\left<cc\right>$ and the global transitivity $T$. The difference between the two definitions is that the average of Eq. (\ref{globalTrans}) gives the same weight to each triangle in the network, whereas Eq. (\ref{avClus}) gives the same weight to each node \cite{Newman2003}. This may lead to slightly different values since nodes with a higher degree may be involved in a higher number of triangles than nodes with a lower degree \cite{F.Costa2007}.

@@

The modularity of a network with $N$ nodes and $L$ edges and partitioned into $n_c$ communities is calculated as follows. Each community has $N_c$ nodes and $L_c$ edges ($c = 1,..., n_c$) \cite{Albert2016}. The modularity of a community, $Q_c$, is calculated by measuring the difference between the network's real wiring diagram (which is given by the adjacency matrix $A_{ij}$) and the expected number of edges between $i$ and $j$ if the network is randomly wired (where $i$ and $j$ are two nodes that lie in the same community $C_c$) \cite{Albert2016}. This last quantity is given by $p_{ij}$, which is obtained by randomizing the original network while keeping the expected degree of each node unchanged \cite{Albert2016}. Hence, the modularity of a community $Q_c$ becomes

@@

The derivations of these formulas can be found in Appendix \ref{WS}, Section \ref{ringclus}. The regular ring lattice thus has high clustering and an average path length that grows linearly with the system size $N$. \\

@@

\textbf{The stochastic Watts-Strogatz block model}\\
